---
title: "Getting Started with hvtiRlifetables"
author: "John Ehrlinger"
date: today
format:
  html:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with hvtiRlifetables}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
```

## What is the dashed line?

Most survival figures from our group carry two curves. The solid one is the
study cohort, a Kaplan-Meier or a parametric fit. The dashed one is what the
same patients could have expected had they never been operated on: the
survival of a US population of the same age, sex and race. A reader's eye goes
straight to the gap between them.

In SAS that dashed line came from the macro `%usmatchd`. `hvtiRlifetables` is
its R replacement. You give it each patient's age, sex and race, a vector of
follow-up times and a reference vintage, and it returns the matched survival
and hazard, per patient or averaged over the cohort.

This vignette walks one small synthetic cohort through that workflow, end to
end. It assumes you know R and survival analysis; it does not assume you have
read the macro.

```{r library}
library(hvtiRlifetables)
```

## Why a fitted model, and not a life table?

You might expect a package like this to carry life tables and interpolate
them. It does not, because `%usmatchd` never did. The macro runs `PROC HAZPRED`
against a stored three-phase hazard fit (Blackstone, Naftel and Turner, 1986)
on the *age* axis, with time zero at birth rather than at surgery. A patient
who is 70 at operation and followed for 5 years is read off that one smooth
curve twice, at 70 and at 75, and the matched survival is the ratio:

$$
S_{matched}(t) = \exp\{-(H(\text{age} + t) - H(\text{age}))\}
$$

So the package ships the fitted parameter blocks, nine strata per vintage, and
evaluates them through `TemporalHazard`. `survival::survexp.usr` is not a
substitute; it is a table, and it does not reproduce the macro's numbers in
any of the vintage and interpolation combinations we tried.

The R side has been checked against production SAS output: all 32 stored
`%usmatchd` answers from two studies match to 7.9e-15 at worst.

## Choosing a vintage

Every call names a reference vintage, and there is no default. Start by
looking at what is available:

```{r vintages}
v <- us_lifetable_vintages()
v[, c("vintage", "n_strata", "nonwhite_code", "added")]
```

The vintages are not interchangeable. They are separately fitted models with
different structure, so the same patient gets a different answer from each,
and the difference is large enough to matter on a figure.

Read the `nonwhite_meaning` column before you name a stratum anywhere a reader
will see it:

```{r nonwhite-meaning}
#| results: asis
for (i in seq_len(nrow(v))) {
  cat("- **", v$vintage[i], "**: ", v$nonwhite_meaning[i], "\n", sep = "")
}
```

The code `b` in `table2023` does **not** mean Black. It is a risk-weighted
average of Black, Asian, American Indian and Hispanic death rates, stored under
`b` to keep the macro's naming. The macro's own comment says otherwise and is
wrong. A figure legend or a methods sentence that says "Black" because the code
says `b` is a clinically wrong statement, and one that looks exactly like what
everyone expects. Take the wording from `nonwhite_meaning`, never from the code.

Leaving the vintage out is an error, on purpose:

```{r no-vintage}
#| error: true
us_matched(age = 70, male = 1, other = 0, times = 5)
```

The macro's default moved twice, from `table84` to `table2008` and then to
`table2023`, and jobs re-run across either move got different numbers with no
warning. Writing the vintage into the analysis code is the only way a reader
can tell which reference population a figure used.

## A small cohort

The rest of this vignette uses 60 synthetic patients. Nothing here comes from a
study; the ages and indicators are drawn at random.

```{r cohort}
set.seed(1986)
n <- 60
cohort <- data.frame(
  id    = sprintf("pt%02d", seq_len(n)),
  age   = round(stats::runif(n, 45, 85), 1),
  male  = stats::rbinom(n, 1, 0.6),
  other = stats::rbinom(n, 1, 0.2)
)
head(cohort)
```

`male` and `other` use the macro's coding unchanged: `male` is 1 for male and
0 for female, `other` is 1 for the vintage's non-white category and 0 for
white. Age is always in years.

## One curve per patient

`us_matched()` takes the cohort's columns as vectors, along with the follow-up
times at which you want the curve:

```{r matched}
x <- us_matched(
  age     = cohort$age,
  male    = cohort$male,
  other   = cohort$other,
  times   = 0:10,
  id      = cohort$id,
  vintage = "table2023"
)
head(x, 4)
```

The result has one row per patient per time. Three columns carry the answer:

- `agesurv` is survival from birth to the patient's current age. It is
  constant within a patient.
- `smatched` is the reference survival over follow-up, conditional on having
  reached that age, so it starts at 1.
- `hmatched` is the reference hazard at each time.

`table = "sexrace"` is the default and matches on all four sex and race
crossings. `"race"`, `"sex"` and `"overall"` match more coarsely, mirroring the
macro's `TABLE=` option.

## The cohort curve

A figure does not want 60 dashed lines. It wants one, the cohort's expected
survival at each time, and `us_cohort_curve()` gives you that:

```{r cohort-curve}
expected <- us_cohort_curve(x)
expected
```

The averaging is the macro's, and it is made in this one function so that two
studies cannot average differently without noticing. Survival and the density
(`hmatched * smatched`) are averaged across patients at each time, unweighted,
and the mean hazard is recovered by dividing one by the other. The `hmatched`
column is therefore **not** the mean of the individual hazards; in a mixed-age
cohort the two differ. `us_matched(..., individual = FALSE)` returns the same
curve in one step.

## Reporting by group

Most papers break the comparison down: by age band, by treatment arm, by era.
`us_cohort_curve()` takes a `by` argument for that. Because `x` has one row
per patient per time, a patient-level grouping is expanded to it by `id`:

```{r by-group}
band <- cut(cohort$age, breaks = c(0, 70, Inf), right = FALSE,
            labels = c("under 70", "70 and over"))
by_band <- us_cohort_curve(x, by = band[match(x$id, cohort$id)])
by_band[by_band$time %in% c(0, 5, 10), ]
```

Keep `by` and `table` apart in your head. `table` decides which strata the
reference *life table* is built from. `by` decides how the resulting curves
are grouped for *reporting*. A report by age band is normally still matched on
`"sexrace"`; the band only changes which patients each dashed line averages
over.

Plotted, these are the two dashed lines you would overlay on each band's
actuarial curve:

```{r by-group-plot}
#| fig-alt: >-
#|   Expected US-matched survival over ten years for two synthetic age bands.
#|   The 70 and over band falls to about 0.5, the under 70 band to about 0.87.
plot(NULL, xlim = c(0, 10), ylim = c(0, 1),
     xlab = "Years after operation", ylab = "Expected survival")
for (g in levels(band)) {
  d <- by_band[by_band$group == g, ]
  lines(d$time, d$smatched, lty = 2, lwd = 2,
        col = if (g == "under 70") "grey30" else "grey60")
}
legend("bottomleft", legend = levels(band), lty = 2, lwd = 2,
       col = c("grey30", "grey60"), bty = "n")
```

## How much does the vintage matter?

The same cohort against all three vintages, ten years out:

```{r vintage-compare}
at_10 <- vapply(v$vintage, function(vin) {
  curve <- us_matched(cohort$age, cohort$male, cohort$other,
                      times = c(0, 10), vintage = vin, individual = FALSE)
  curve$smatched[curve$time == 10]
}, numeric(1))
round(at_10, 3)
```

For this cohort, `table84` and `table2023` are nearly nine percentage points
apart at ten years. The closest pair, `table2008` and `table2023`, still differ
by close to three, which is small enough to look plausible on a figure and
survive review. That is why the package makes you say which one you meant.

## Three things to know before you rely on it

**`hmatched` is per year, always.** Whatever `scale` you pass (`"years"`,
`"months"` or `"days"`), `time` is in those units but `hmatched` stays a
hazard per year. That reproduces the macro's code, not its header comment,
which claims otherwise. If you need a per-month hazard, divide by 12 yourself.

**The fits stop at age 110.** If any patient's age plus follow-up passes 110,
`us_matched()` stops rather than extrapolating.

**A curve that does not move is an error.** If the matched survival comes back
identically 1 or 0 across the requested times, the function stops and says so.
That failure has turned up often enough in studies to be worth catching before
it reaches a figure.

## Where to go next

- `?us_matched` documents every argument, including `scale`, `table` and the
  output columns.
- `?us_cohort_curve` covers the averaging and the forms `by` accepts,
  including a named list for crossed groupings.
- `?us_lifetable_model` returns the raw fitted parameters for one vintage and
  stratum, for when a number looks wrong and you want to see the model behind
  it.
- `?us_lifetable_models` describes the shipped dataset and its `_STATUS_` gate.
